---
title: "1800s emigration to America"
author: "Oliver Hofman"
date: "2026-08-03"
output: html_document
---


```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidytext)
library(ggwordcloud)
library(syuzhet)
library (stopwords)
library(ggplot2)
library(dplyr)
library(readr)

```

# Loading the data into R
using Swagger UI, a link is created from the search "(Amerika OR USA OR "de forenede Stater") AND (udvandre* OR emigr*) AND py:[1840 TO 1880] NOT (Thingvalla OR "Allan Line" OR Cunard)"
```{r Loading Data}
link <-"https://labs.statsbiblioteket.dk/labsapi/api/aviser/export/fields?query=%28Amerika%20OR%20USA%20OR%20%22de%20forenede%20Stater%22%29%20AND%20%28udvandre%2A%20OR%20emigr%2A%29%20AND%20py%3A%5B1840%20TO%201880%5D%20NOT%20%28Thingvalla%20OR%20%22Allan%20Line%22%20OR%20Cunard%29&fields=link&fields=recordID&fields=timestamp&fields=pwa&fields=cer&fields=fulltext_org&fields=pageUUID&fields=editionUUID&fields=titleUUID&fields=editionId&fields=familyId&fields=newspaper_page&fields=newspaper_edition&fields=lplace&fields=location_name&fields=location_coordinates&max=-1&structure=header&structure=content&format=CSV"
  
URLdecode(link)

NewsP_1800s <- read.csv(link)
```

# Analysing the datas origin
Here its seen that Copenhagen is in the clear lead, with quite the drop to the next most. 
it also shows that "Dagstelegraphen" is the paper with most mentions.
```{r Origin of the data}
NewsP_1800s %>% 
  count(lplace, sort = TRUE)

NewsP_1800s %>% 
  count(familyId, sort = TRUE)
```

#Textmining
Here the articles are divided up into singular words and then edited by using 2 lists of stopwords pertaining to  1800s danish and modern danish

```{r Textmining}

# Tokenize the newspaper text
NewsP_1800s_tidy <- NewsP_1800s %>%
  unnest_tokens(word, fulltext_org)

# Read the 1800s stopword list
stopwords_1800s <- read_csv(
  "https://gist.githubusercontent.com/maxodsbjerg/1537cf14c3d46b3d30caa5d99f8758e9/raw/9f044a38505334f035be111c9a3f654a24418f6d/stopord_18_clean.csv",
  show_col_types = FALSE
)

# Load modern Danish stopwords
stopwords_modern <- tibble(
  word = stopwords::stopwords("da")
)

# Combine the two stopword lists and remove duplicates
stopwords_all <- bind_rows(
  stopwords_1800s,
  stopwords_modern
) %>%
  distinct(word)

# Remove all stopwords and count the remaining words
NewsP_1800s_wStopwords <- NewsP_1800s_tidy %>%
  anti_join(stopwords_all, by = "word") %>%
  count(word, sort = TRUE)

```

#Removing numerals and creating a top 100
Here numbers are filtered out, as to only get actual words that can be analysed. Then its put into a top 100 of most mentions
```{r Removing numerals and creating a top 100}
NewsP_1800s_nonum <- NewsP_1800s_wStopwords %>% 
filter(!str_detect(word, "[0-9]"))

 NewsP_1800s_top100 <- head(NewsP_1800s_nonum, 100)
 
 NewsP_1800s_top100
```

#Wordcloud
by creating a wordcloud with the top 100, its made easier to visualize whats mentioned the most.

```{r}
ggplot(data = NewsP_1800s_top100, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n), shape = "star") +
  scale_size_area(max_size = 12) +
  scale_color_gradientn(colors = c("darkgreen","blue","red")) +
  theme_minimal()
```

#Sentiment analysis
here the words get joined together back into the whole article, as to get a overall sentiment score from the articles.

```{r}
fulltext_tidy_recordID <- NewsP_1800s_tidy %>% 
  anti_join(stopwords_1800s, by = c("word" = "word")) %>% 
  group_by(recordID) %>%  
  summarise(fulltext_clean = paste(word, collapse = " "), .groups = "drop")

sentiment_text_recordID <- fulltext_tidy_recordID %>% 
  mutate(
    fulltext_sentiment = map_dbl(
      fulltext_clean,
      function(x) get_sentiment(x, method = "afinn")
    )
  )
sentiment_by_record_1800s <- sentiment_text_recordID %>%
  group_by(recordID) %>%
  summarise(
    mean_sentiment = mean(fulltext_sentiment, na.rm = TRUE),
    n = n()
  )


```

#Visualizing the Sentiment
here the articles sentiment score is plotted into a histogram to better visualize the spread of sentiment across all the articles

```{r}
library(ggplot2)

ggplot(sentiment_by_record_1800s,
       aes(x = mean_sentiment)) +
  geom_histogram(bins = 30,
                 fill = "Blue",
                 color = "white") +
  theme_minimal() +
  labs(
    title = "Fordeling af sentiment",
    x = "Sentiment score",
    y = "Antal tekster"
  )
ggsave("histogram 1800s.png", width = 8, height = 6, dpi = 300)
```

#Average Sentiment Score
here a average of the sentiment is calculated 

```{r}
mean(sentiment_by_record_1800s$mean_sentiment,
     na.rm = TRUE)

```

#Categorizing the sentiment
by dividing all the articles into 3 Categories, "Positiv", "Negativ" and "Neutral" and counting how many of the articles land into each categories 

```{r}
sentiment_CAT_1800s <- sentiment_by_record_1800s %>%
  mutate(sentiment_category = case_when(
    mean_sentiment > 0 ~ "Positiv",
    mean_sentiment < 0 ~ "Negativ",
    TRUE ~ "Neutral"
  ))

sentiment_CAT_1800s %>%
  count(sentiment_category)

```