---
title: "1860s Sentiment towards animal cruelty"
author: "Oliver Hofman"
date: "2026-05-20"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidytext)
library(ggwordcloud)
library(syuzhet)

```

# Loading the data into R
using Swagger UI, a link is created from the search "dyrplager* py:[1860-69]" in Mediestream
```{r}
link <- "https://labs.statsbiblioteket.dk/labsapi/api/aviser/export/fields?query=dyrplager%2A%20py%3A%5B1860%20TO%201869%5D&fields=link&fields=recordID&fields=timestamp&fields=pwa&fields=cer&fields=fulltext_org&fields=pageUUID&fields=editionUUID&fields=titleUUID&fields=editionId&fields=familyId&fields=newspaper_page&fields=newspaper_edition&fields=lplace&fields=location_name&fields=location_coordinates&max=-1&structure=header&structure=content&format=CSV"

URLdecode(link)

Dyr_1860s <- read.csv(link)
```

# Analysing the datas origin
Here its seen that Copenhagen is in the clear lead, with quite the drop to the next most. 
it also shows that "flyveposten" is the paper with most mentions, with berlingske right behind it.
```{r}
Dyr_1860s %>% 
  count(lplace, sort = TRUE)

Dyr_1860s %>% 
  count(familyId, sort = TRUE)
```

#Textmining
Here the articles are divided up into singular words and then edited by using a list of stopwords pertaining to especially 1800s danish.

```{r}
Dyr_1860s_tidy <- Dyr_1860s %>% 
  unnest_tokens(word, fulltext_org)

stopwords_1800s <- read_csv("https://gist.githubusercontent.com/maxodsbjerg/1537cf14c3d46b3d30caa5d99f8758e9/raw/9f044a38505334f035be111c9a3f654a24418f6d/stopord_18_clean.csv")

Dyr_1860s_wStopwords<- Dyr_1860s_tidy %>% 
  anti_join(stopwords_1800s) %>% 
  count(word, sort = TRUE)
```

#Removing numerals and creating a top 100
Here numbers are filtered out, as to only get actual words that can be analysed. Then its put into a top 100 of most mentions
```{r}
Dyr_1860s_nonum <- Dyr_1860s_wStopwords %>% 
filter(!str_detect(word, "[0-9]"))

 Dyr_1860s_top100 <- head(Dyr_1860s_nonum, 100)
 
 Dyr_1860s_top100
```

#Wordcloud
by creating a wordcloud with the top 100, its made easier to visualize whats mentioned the most.

```{r}
ggplot(data = Dyr_1860s_top100, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n), shape = "star") +
  scale_size_area(max_size = 12) +
  scale_color_gradientn(colors = c("darkgreen","blue","red")) +
  theme_minimal()

```

#Sentiment analysis
here the text a once again edited with the stopwords, and then joined together as to get a overall sentiment score from the articles.

```{r}
fulltext_tidy_recordID <- Dyr_1860s_tidy %>% 
  anti_join(stopwords_1800s, by = c("word" = "word")) %>% 
  group_by(recordID) %>%  
  summarise(fulltext_clean = paste(word, collapse = " "), .groups = "drop")

sentiment_text_recordID <- fulltext_tidy_recordID %>% 
  mutate(
    fulltext_sentiment = map_dbl(
      fulltext_clean,
      function(x) get_sentiment(x, method = "afinn")
    )
  )
sentiment_by_record_1860s <- sentiment_text_recordID %>%
  group_by(recordID) %>%
  summarise(
    mean_sentiment = mean(fulltext_sentiment, na.rm = TRUE),
    n = n()
  )


```

#Visualizing the Sentiment
here the articles sentiment score is plotted into a histogram to better visualize the spread of sentiment across all the articles

```{r}
library(ggplot2)

ggplot(sentiment_by_record_1860s,
       aes(x = mean_sentiment)) +
  geom_histogram(bins = 30,
                 fill = "Blue",
                 color = "white") +
  theme_minimal() +
  labs(
    title = "Fordeling af sentiment",
    x = "Sentiment score",
    y = "Antal tekster"
  )
```

#Average Sentiment Score
here a average of the sentiment is calculated 

```{r}
mean(sentiment_by_record_1860s$mean_sentiment,
     na.rm = TRUE)

```

#Categorizing the sentiment
by dividing all the articles into 3 Categories, "Positiv", "Negativ" and "Neutral" and counting how many of the articles land into each categories 

```{r}
sentiment_CAT_1860s <- sentiment_by_record_1860s %>%
  mutate(sentiment_category = case_when(
    mean_sentiment > 0 ~ "Positiv",
    mean_sentiment < 0 ~ "Negativ",
    TRUE ~ "Neutral"
  ))

sentiment_CAT_1860s %>%
  count(sentiment_category)
```
