---
title: Game of Thrones - 'Text mining, sentiment analysis, and visualization'
date: '27-3-2026'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

library(tidyverse)
library(here)

# For text mining:

library(pdftools)
library(tidytext)
library(textdata) 
library(ggwordcloud)
```



### Get the GoT Script:
```{r get-document}
GoT_path <- here("data","GoT.pdf")
GoT_path
GoT_text <- pdf_text(GoT_path)
GoT_text
```

Some things to notice:

Example: Just want to get text from a single page (e.g. Page 30)? 
```{r single-page}
GoT_p45 <- GoT_text[45]
GoT_p45
```


### Some wrangling:

- Split up pages into separate lines (separated by `\n`) using `stringr::str_split()`
- Unnest into regular columns using `tidyr::unnest()`
- Remove leading/trailing white space with `stringr::str_trim()`

```{r split-lines}
GoT_df <- data.frame(GoT_text) %>% 
  mutate(text_full = str_split(GoT_text, pattern = '\n')) %>% 
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full)) 
GoT_df

# More information: https://cran.r-project.org/web/packages/stringr/vignettes/regular-expressions.html
```

Now each line, on each page, is its own row, with extra starting & trailing spaces removed. 

### Get the tokens (individual words) in tidy format

Use `tidytext::unnest_tokens()` (which pulls from the `tokenizer`) package, to split columns into tokens. We are interested in *words*, so that's the token we'll use:

```{r tokenize}
GoT_tokens <- GoT_df %>% 
  unnest_tokens(word, text_full)
GoT_tokens



# See how this differs from `GoT_df`
# Each word has its own row!
```

Let's count the words!
```{r count-words}
GoT_wc <- GoT_tokens %>% 
  count(word) %>% 
  arrange(-n)
GoT_wc
```

OK...so we notice that a whole bunch of things show up frequently that we might not be interested in ("a", "the", "and", etc.). These are called *stop words*. Let's remove them. 

### Remove stop words:

See `?stop_words` and `View(stop_words)`to look at documentation for stop words lexicons.

We will *remove* stop words using `tidyr::anti_join()`:
```{r stopwords}
?stop_words
GoT_stop <- GoT_tokens %>% 
  anti_join(stop_words) %>% 
  select(-GoT_text)
```

Now check the counts again: 
```{r count-words2}
GoT_swc <- GoT_stop %>% 
  count(word) %>% 
  arrange(-n)
GoT_swc
```

What if we want to get rid of all the numbers (non-text) in `GoT_stop`?
```{r skip-numbers}
# This code will filter out numbers by asking:
# If you convert to as.numeric, is it NA (meaning those words)?
# If it IS NA (is.na), then keep it (so all words are kept)
# Anything that is converted to a number is removed

GoT_no_numeric <- GoT_stop %>% 
  filter(is.na(as.numeric(word)))

GoT_no_numeric %>% 
  count(word) %>% 
  arrange(-n)
```

### A word cloud of GoT report words (non-numeric)

See more: https://cran.r-project.org/web/packages/ggwordcloud/vignettes/ggwordcloud.html

```{r wordcloud-prep}
# There are almost 11209 unique words 
length(unique(GoT_no_numeric$word))

# We probably don't want to include them all in a word cloud. Let's filter to only include the top 100 most frequent?
GoT_top100 <- GoT_no_numeric %>% 
  count(word) %>% 
  arrange(-n) %>% 
  head(100)

GoT_top100
```

```{r wordcloud}
GoT_cloud <- ggplot(data = GoT_top100, aes(label = word)) +
  geom_text_wordcloud() +
  theme_minimal()


GoT_cloud
```

That's underwhelming. Let's customize it a bit:
```{r wordcloud-pro}
ggplot(data = GoT_top100, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n), shape = "star") +
  scale_size_area(max_size = 12) +
  scale_color_gradientn(colors = c("darkgreen","blue","red")) +
  theme_minimal()

ggsave("starGoT.png", width = 10)
```


### Sentiment analysis

“The three general-purpose lexicons are

  -  AFINN from Finn Årup Nielsen,
  -  bing from Bing Liu and collaborators, and
  -  nrc from Saif Mohammad and Peter Turney

All three of these lexicons are based on unigrams, i.e., single words. These lexicons contain many English words and the words are assigned scores for positive/negative sentiment, and also possibly emotions like joy, anger, sadness, and so forth. The AFINN lexicon assigns words with a score that runs between -5 and 5, with negative scores indicating negative sentiment and positive scores indicating positive sentiment.  The bing lexicon categorizes words in a binary fashion into positive and negative categories. The nrc lexicon categorizes words in a binary fashion (“yes”/“no”) into categories of positive, negative, anger, anticipation, disgust, fear, joy, sadness, surprise, and trust.  All of this information is tabulated in the sentiments dataset, and tidytext provides a function get_sentiments() to get specific sentiment lexicons without the columns that are not used in that lexicon."

Let's explore the sentiment lexicons. "bing" is included, other lexicons ("afinn", "nrc", "loughran") you'll be prompted to download.

**WARNING:** These collections include very offensive words. I urge you to not look at them in class.

"afinn": Words ranked from -5 (very negative) to +5 (very positive)
```{r afinn}
get_sentiments(lexicon = "afinn")
# You should be prompted to install lexicon - choose yes!


# Let's look at the pretty positive words:
afinn_pos <- get_sentiments("afinn") %>% 
  filter(value %in% c(3,4,5))

# Do not look at negative words in class. 
afinn_pos
```

bing: binary, "positive" or "negative"
```{r bing}
get_sentiments(lexicon = "bing")
# You should be prompted to install lexicon - choose yes!

```


Now nrc:
```{r nrc}
get_sentiments(lexicon = "nrc")
# You should be prompted to install lexicon - choose yes!

```

Let's do sentiment analysis on the GoT text data using afinn, and nrc. 


### Sentiment analysis with afinn: 

First, bind words in `GoT_stop` to `afinn` lexicon:
```{r bind-afinn}
GoT_afinn <- GoT_stop %>% 
  inner_join(get_sentiments("afinn"))
GoT_afinn
```

Let's find some counts (by sentiment ranking):
```{r count-afinn}
GoT_afinn_hist <- GoT_afinn %>% 
  count(value)

# Plot them: 
ggplot(data = GoT_afinn_hist, aes(x = value, y = n)) +
  geom_col(aes(fill = value)) +
  theme_bw()
```

Investigate some of the words in a bit more depth:
```{r afinn-2}
# What are these '-2' words?
GoT_afinn2 <- GoT_afinn %>% 
  filter(value == -2)
```

```{r afinn-2-more}
# Check the unique 2-score words:
unique(GoT_afinn2$word)

# Count & plot them
GoT_afinn2_n <- GoT_afinn2 %>% 
  count(word, sort = TRUE) %>% 
  mutate(word = fct_reorder(factor(word), n))


ggplot(data = GoT_afinn2_n, aes(x = word, y = n)) +
  geom_col() +
  coord_flip() +
  theme_bw()
ggsave("sentiment_plotGoT.png", width = 10, height = 20,limitsize = FALSE)
# OK so what's the deal with confidence? And is it really "positive" in the emotion sense? 
```



Lets summarize sentiment for the script: 
```{r summarize-afinn}
GoT_summary <- GoT_afinn %>% 
  summarize(
    mean_score = mean(value),
    median_score = median(value)
  )
GoT_summary 
```

The mean and median indicate *slightly* negative overall sentiments based on the AFINN lexicon. 

### NRC lexicon for sentiment analysis

We can use the NRC lexicon to start "binning" text by the feelings they're typically associated with. As above, we'll use inner_join() to combine the GoT non-stopword text with the nrc lexicon: 

```{r bind-bing}
GoT_nrc <- GoT_stop %>% 
  inner_join(get_sentiments("nrc"))
GoT_nrc
```

Wait, won't that exclude some of the words in our text? YES! We should check which are excluded using `anti_join()`:

```{r check-exclusions}
GoT_exclude <- GoT_stop %>% 
  anti_join(get_sentiments("nrc"))


# View(GoT_exclude)

# Count to find the most excluded:
GoT_exclude_n <- GoT_exclude %>% 
  count(word, sort = TRUE)

head(GoT_exclude_n)
```

**Lesson: always check which words are EXCLUDED in sentiment analysis using a pre-built lexicon! **

Now find some counts: 
```{r count-bing}
GoT_nrc_n <- GoT_nrc %>% 
  count(sentiment, sort = TRUE)

# And plot them:

ggplot(data = GoT_nrc_n, aes(x = sentiment, y = n)) +
  geom_col(aes(fill = sentiment))+
  theme_bw()
```

Or count by sentiment *and* word, then facet:
```{r count-nrc}
GoT_nrc_n5 <- GoT_nrc %>% 
  count(word,sentiment, sort = TRUE) %>% 
  group_by(sentiment) %>% 
  top_n(5) %>% 
  ungroup()

GoT_nrc_gg <- ggplot(data = GoT_nrc_n5, aes(x = reorder(word,n), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, ncol = 2, scales = "free") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Word", y = "count")

# Show it
GoT_nrc_gg

# Save it
ggsave(plot = GoT_nrc_gg, 
       here("figures","GoT_nrc_sentiment.png"), 
       height = 8, 
       width = 5)

```

Wait, so "lord" is showing up in NRC lexicon as "negative" and "positive"? Let's check:
```{r nrc-confidence}
conf <- get_sentiments(lexicon = "nrc") %>% 
  filter(word == "lord")

# Yep, check it out:
conf
```


I went in expecting a lot of negative emotions, since the book is about feudal politics and war, but was surprised
to find that there were a lot more mixed emotions. although the median and average scores from the summary shows it
still is slightly negative. A term that shows up like "confidence" did, was the word "Lord". showing up both as
disgust, yet also trust and positive. "Stark" shows up a lot, but i think it has lumped one of the characters
names in with the other words, so it should be disregarded, along with other names, maybe even Lord itself. as it
serves mostly as a part of someones extended name throughout the book.
